---
title: "SURV727 HW#5"
author: "Yael Beshaw"
date: "2024-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the Necessary Packages
```{r}
library(censusapi)
library(tidyverse)
library(magrittr)
library(factoextra)
```

# Exploring ACS Data
```{r}
cs_key <- "4dfcefb5de015bd6b5ce2762c614a779762b6c0b"

#Illinois DF with county, pop, hh_income, and income
acs_il_c <- getCensus(name = "acs/acs5",
                      vintage = 2016,
                      vars = c("NAME", "B01003_001E",
                               "B19013_001E", "B19301_001E"),
                      region = "county:*",
                      regionin = "state:17",
                      key= cs_key) %>%
  rename(pop = B01003_001E,
         hh_income = B19013_001E,
         income = B19301_001E)
head(acs_il_c)

#Pull map data for Illinois into a data frame
il_map <- map_data("county", region = "illinois")
head(il_map)
```
Join the ACS data with the map data. Note that il_map has a column subregion 
which includes county names. We need a corresponding variable in the ACS data to
join both data sets. This needs some transformations, among which the function
tolower() might be useful. Call the joined data acs_map. 

```{r}
#acs_il_c$NAME has "______ County, Illinois", I have to get rid of "County,
#Illinois in order to completely match it to il_map$subregion and then lowercase

acs_il_c$NAME <- gsub(" County, Illinois", "", acs_il_c$NAME) #remove extra words
acs_il_c$NAME <- tolower(acs_il_c$NAME) #lowercase

il_map$subregion <- tolower(il_map$subregion) #make sure this is lc as well

acs_map <- left_join(il_map, acs_il_c, by = c("subregion" = "NAME")) #join

#double check everything worked
head(acs_map)
```

Plot a map og Illinois with Counties colored by per capita income
```{r}
ggplot(acs_map) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = income))
```
# Hierarchial Clustering
We want to find clusters of counties that are similar in their population, 
average household income and per capita income. First, clean the data so that
you have the appropriate variables to use for clustering. Next, create the 
distance matrix of the cleaned data. This distance matrix can be used to cluster
counties, e.g. using the ward method.

```{r}
#appropriate variables 
# population (pop)
# average household income (hh_income)
# per capita income (income)

#clean the data so that you have the appropriate variables to use for clustering
acs_heir <- acs_il_c %>%
  select(pop, hh_income, income)%>%
  mutate_all(scale)%>%
  drop_na()%>%
  rename(pop = pop,
         hh_income = hh_income,
         income = income)

#create the distance matrix of the cleaned data
distance_matrix <- dist(acs_heir)
```

Plot the dendrogram to find a reasonable number of clusters. Draw boxes around 
the clusters of your cluster solution. 
```{r}
#Using the Ward Method
acs_ward <- hclust(distance_matrix, method = "ward.D2")
plot(acs_ward, main = "Ward", xlab = "", sub = "")

rect.hclust(acs_ward, 
            k = 10, 
            border = "red")
```
Visualize the county clusters on a map. For this task, create a new acs_map 
object that now also includes cluster membership as a new column. This column 
should be called cluster.

```{r}
#create a new acs_map object that now also includes cluster membership 
#as a new column
cluster_membership <- cutree(acs_ward, 10)
acs_il_c$cluster <- cluster_membership
acs_map <- left_join(il_map, acs_il_c, by = c("subregion" = "NAME"))

#acs_map$cluster #check

ggplot(acs_map) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(cluster)))
```

# Census Tracts
For the next section we need ACS data on a census tract level. We use the same
variables as before.
```{r}
acs_il_t <- getCensus(name = "acs/acs5", vintage = 2016, 
                      vars = c("NAME", 
                               "B01003_001E", 
                               "B19013_001E", 
                               "B19301_001E"), 
                      region = "tract:*", 
                      regionin = "state:17", key = cs_key) %>% 
  mutate_all(funs(ifelse(.==-666666666, NA, .))) %>%
  rename(pop = B01003_001E,
         hh_income = B19013_001E,
         income = B19301_001E)

head(acs_il_t)
```


# k-Means

As before, clean our data for clustering census tracts based on population, 
average household income and per capita income
```{r}
acs_new <- acs_il_t %>%
  select(pop, hh_income, income) %>%
  mutate_all(scale) %>%
  drop_na()
```

Since we want to use K Means in this section, we start by determining the 
optimal number of K that results in Clusters with low within but high between 
variation. Plot within cluster sums of squares for a range of K (e.g. up to 20).
```{r}
fviz_nbclust(acs_new, #data set we want to use
             kmeans, #cluster method
             method = "wss", #method used for est the optimal number of clusters
             k.max = 20) #5 seems to be the optimal number of clusters
```

Run kmeans() for the optimal number of clusters based on the plot above.
```{r}
km <- kmeans(acs_new, 5, nstart = 20)
```

Find the mean population, household income and per capita income grouped by 
clusters.
```{r}
acs_new$cluster <- km$cluster

means <- acs_new %>%
  group_by(cluster) %>%
  summarise(
    mean_pop = mean(pop, na.rm = TRUE),
    mean_hh_income = mean(hh_income, na.rm = TRUE),
    mean_income = mean(income, na.rm = TRUE)
  )
means
```

In addition, display the most frequent county that can be observed 
within each cluster.
```{r}
acs_il_t_clean <- acs_il_t %>%
  filter(complete.cases(pop, hh_income, income)) #make sure that the dim match
acs_il_t_clean$cluster <- km$cluster #add km$cluser to the original and clean df

#find most freq.county
frequent_county <- acs_il_t_clean %>%
  group_by(cluster) %>%
  summarise(
    most_frequent_county = names(sort(table(NAME), decreasing = TRUE))[1]
  )

frequent_county
```

As you might have seen earlier, it’s not always clear which number of clusters
is the optimal choice. To automate K Means clustering, program a function based
on kmeans() that takes K as an argument. You can fix the other arguments, e.g. 
such that a specific dataset is always used when calling the function.
```{r}
automatic_kmeans <- function (df, k){
  cluster <- kmeans(df, k, nstart = 20)
  return(cluster$cluster)
}

#automatic_kmeans(acs_new,6 )
```


We want to utilize this function to iterate over multiple Ks
(e.g., K = 2, ..., 10) and – each time– add the resulting cluster membership as
a new variable to our (cleaned) original data frame (acs_il_t). There are 
multiple solutions for this task, e.g. think about the apply family or
for loops.
```{r}

automatic_kmeans <- function (df, k){
  if (k >= nrow(df)) { #3109 or higher in the case of acs_new
    print("error: more cluster centers than distinct data points")
    return(NULL) #set as NULL to avoid errors when this occurs 
  } 
  cluster <- kmeans(df, k, nstart = 20)
  return(cluster$cluster)
}


for (k in 2:10) {
  cluster_loop <- automatic_kmeans(acs_new, k)
  if (!is.null(cluster_loop)) { #if not NULL (bigger than data points), proceed
    acs_il_t_clean[[paste0("cluster_", k)]] <- cluster_loop
  }
}
```


Finally, display the first rows of the updated data set (with multiple
cluster columns)
```{r}
head(acs_il_t_clean)
```



