---
title: 'Assignment #4 SURV727'
output:
  html_document:
    df_print: paged
---

## Setup and Packages
```{r}
library(tidyverse)
library(DBI)
library(dbplyr)
library(bigrquery)
```

## Database Connection
```{r}
project <- "surv727-hw4-2024"

con <- dbConnect(
  bigrquery::bigquery(),
  project = "bigquery-public-data",
  dataset = "chicago_crime",
  billing = project
  )
con

```

Observe available tables
```{r}
dbListTables(con)
```

## Creating Queries
#Using SQL
```{r}
#check the column names
dbListFields(con, "crime")
```

Write a first query that counts the number of rows of the ‘crime‘ table in the
year 2016. Use code chunks with {sql connection = con} in order to write SQL 
code within the document.

This query indicates to us that the number of rows in this table is 269,922.
```{sql connection = con}

SELECT count(primary_type), count(*)
FROM crime
WHERE year = 2016
LIMIT 10;
```


Next, count the number of arrests grouped by primary_type in 2016. Note that is
a somewhat similar task as above, with some adjustments on which rows should be 
considered. Sort the results, i.e. list the number of arrests in a descending
order.

The results indicate that Narcotics is the type with the greatest number of 
arrests, with Battery following close behind.

```{sql connection = con}
SELECT primary_type, count(*) AS num_arrests
FROM crime
WHERE year = 2016 and arrest= TRUE
GROUP BY primary_type
ORDER BY num_arrests DESC 
LIMIT 10;
```

We can also use the date for grouping. Count the number of arrests grouped by
hour of the day in 2016. You can extract the latter information from date via 
EXTRACT(HOUR FROM date). Which time of the day is associated with the most 
arrests?

The 19th hour (aka 7pm) is associated with the most arrests. The top four
hours with the most arrests are between 6pm and 9pm which indicates that
evenings are when the most arrests happen in Chicago.

```{sql connection = con}

SELECT EXTRACT(HOUR FROM date) AS hour_of_day, count(*) AS num_arrest
FROM crime
WHERE year = 2016 and arrest= TRUE
GROUP BY hour_of_day
ORDER BY num_arrest DESC
LIMIT 10;
```
Focus only on HOMICIDE and count the number of arrests for this incident type, 
grouped by year. List the results in descending order.

Based on these results 2001-2003 are the years with the greatest number of
homicide arrests with 2020 and 2022 following behind.
```{sql connection = con}
SELECT year, count(*) AS h_arrests
FROM crime
WHERE primary_type= "HOMICIDE" and arrest= TRUE
GROUP BY year
ORDER BY h_arrests DESC 
LIMIT 10;
```
Find out which districts have the highest numbers of arrests in 2015 and 2016. 
That is, count the number of arrests in 2015 and 2016, grouped by year and 
district. List the results in descending order.

District 11 has the highest number of arrests in both 2015 and 2016 while 
District 7 has the top 10 ten highest number of arrests in both years 
(#3 and #9).
```{sql connection = con}
SELECT year, district, count(*) AS highest_arrests
FROM crime
WHERE arrest= TRUE and (year = 2015 OR year = 2016)
GROUP BY district, year
ORDER BY highest_arrests DESC 
LIMIT 10;
```
#Using R via DBI package
Lets switch to writing queries from within R via the DBI package. Create a query
object that counts the number of arrests grouped by primary_type of district 11
in year 2016. The results should be displayed in descending order.

For District 11, in 2016 the greatest type of arrests were Narcotics-related.
```{r}
sql <- "SELECT primary_type, count(*) as district_11_2016 
FROM crime 
WHERE arrest= TRUE AND district = 11 AND year= 2016
GROUP BY primary_type
ORDER BY district_11_2016 DESC"

dbGetQuery(con, sql)

```
#Using R via dbplyr
Try to write the very same query, now using the dbplyr package. For this, you 
need to first map the crime table to a tibble object in R.
```{r}
crime <- tbl(con, "crime")
str(crime)
class(crime)
```
Again, count the number of arrests grouped by primary_type of district 11 in 
year 2016, now using dplyr syntax.

The results of utilizing R and dplyr produce what we see in our use of DBI.
```{r}
arrests_byprimary_dist11_2016 <- crime %>%
  select(primary_type, arrest, district, year) %>%
  filter(year == 2016 & district == 11 & arrest== TRUE) %>%
  group_by(primary_type) %>%
  summarise(total_arrests_dist11_2016 = n())%>%
  arrange(desc(total_arrests_dist11_2016))

arrests_byprimary_dist11_2016  
```
Count the number of arrests grouped by primary_type and year, still only for 
district 11. Arrange the result by year.
```{r}
arrests_byprimary_year_dist11 <- crime %>%
  select(primary_type, arrest, district, year) %>%
  filter(district == 11 & arrest== TRUE) %>%
  group_by(primary_type, year) %>%
  summarise(total_arrests_dist11 = n())%>%
  arrange(desc(year))

arrests_byprimary_year_dist11 
```
Assign the results of the query above to a local R object.
```{r}
arrests_byprimary_year_dist11 <- crime %>%
  select(primary_type, arrest, district, year) %>%
  filter(district == 11 & arrest== TRUE) %>%
  group_by(primary_type, year) %>%
  summarise(total_arrests_dist11 = n())%>%
  arrange(desc(year))%>%
  collect()

arrests_byprimary_year_dist11
```

Confirm that you pulled the data to the local environment by displaying the 
first ten rows of the saved data set.
```{r}
str(arrests_byprimary_year_dist11)
head(arrests_byprimary_year_dist11)
```
Close the connection.
```{r}
dbDisconnect(con)
```







