---
title: "SURV727 HW#5"
author: "Yael Beshaw"
date: "2024-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load the Necessary Packages #
```{r}
library(censusapi)
library(tidyverse)
library(magrittr)
library(factoextra)
```

# Exploring ACS Data #
```{r}
cs_key <- "4dfcefb5de015bd6b5ce2762c614a779762b6c0b"

#Illinois DF with county, pop, hh_income, and income
acs_il_c <- getCensus(name = "acs/acs5",
                      vintage = 2016,
                      vars = c("NAME", "B01003_001E",
                               "B19013_001E", "B19301_001E"),
                      region = "county:*",
                      regionin = "state:17",
                      key= cs_key) %>%
  rename(pop = B01003_001E,
         hh_income = B19013_001E,
         income = B19301_001E)
head(acs_il_c)

#Pull map data for Illinois into a data frame
il_map <- map_data("county", region = "illinois")
head(il_map)
```
Join the ACS data with the map data. Note that il_map has a column subregion 
which includes county names. We need a corresponding variable in the ACS data to
join both data sets. This needs some transformations, among which the function
tolower() might be useful. Call the joined data acs_map. 

```{r}
#acs_il_c$NAME has "______ County, Illinois", I have to get rid of "County,
#Illinois in order to completely match it to il_map$subregion and then lowercase

acs_il_c$NAME <- gsub(" County, Illinois", "", acs_il_c$NAME) #remove extra words
acs_il_c$NAME <- tolower(acs_il_c$NAME) #lowercase

il_map$subregion <- tolower(il_map$subregion) #make sure this is lowercase as well

acs_map <- left_join(il_map, acs_il_c, by = c("subregion" = "NAME")) #join

#double check everything worked
head(acs_map)
```

Plot a map og Illinois with Counties colored by per capita income
```{r}
ggplot(acs_map) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = income))
```
# Hierarchial Clustering #
We want to find clusters of counties that are similar in their population, 
average household income and per capita income. First, clean the data so that
you have the appropriate variables to use for clustering. Next, create the 
distance matrix of the cleaned data. This distance matrix can be used to cluster
counties, e.g. using the ward method.

```{r}
#appropriate variables 
# population (pop)
# average household income (hh_income)
# per capita income (income)

#clean the data so that you have the appropriate variables to use for clustering
acs_heir <- acs_map %>%
  select(pop, hh_income, income) %>%
  drop_na() %>%
  mutate_all(scale)
  

#create the distance matrix of the cleaned data
distance_matrix <- dist(acs_heir)
distance_matrix
```
Plot the dendrogram to find a reasonable number of clusters. Draw boxes around 
the clusters of your cluster solution. 
```{r}
#Using the Ward Method
acs_ward <- hclust(distance_matrix, method = "ward.D2")
plot(acs_ward, main = "Ward", xlab = "", sub = "")

rect.hclust(acs_ward, 
            k = 10, 
            border = "red")
```
Visualize the county clusters on a map. For this task, create a new acs_map 
object that now also includes cluster membership as a new column. This column 
should be called cluster.
```{r}
#create a new acs_map object that now also includes cluster membership as a new column
cluster_membership <- cutree(acs_ward, 10)

acs_map$cluster <- cluster_membership
```


